Sampling Strategies in Decisions from Experience

### 2022-02-16 manuscript.Rmd
Review of papaer structure + update of introduction

The human mind is a cognitive system that operates on inputs from an environment or memory.
In the domain of decision making, such inputs provide information about the *properties* of the choice alternatives among which people can choose.
Hereafter, let us treat these alternatives as elements $c$ of the set $C$.
To each $c \in C$ is then assigned some information from the inputs and this information may be used by the mind to learn about the properties of the respective alternative.
Accordingly, decision making can be considered an information processing problem which starts from the set $C$, containing all choice alternatives, and stops at a proper subset $C' \subset C$ with the inclusion map
$$
\begin{aligned}
  \iota : C' &\to C \\
  c &\mapsto c  \quad \forall c \in C'
\end{aligned}
\tag{1}
$$
[@jostMathematicalConcepts2015, p. 15], containing only the chosen alternatives.
Indeed, irrespective of their exact mathematical specification or computational implementation, decision models generally take as inputs information about a set of choice alternatives to return a subset thereof as outputs [@heOntologyDecisionModels2020].

Importantly, information about the choice alternatives can come in many formats.
In *decisions from description* (DfD), inputs take the form of explicit and complete descriptions of all relevant properties of the choice alternatives.
In *decisions from experience* [DfE, @hertwigDecisionsExperienceEffect2004], inputs take the form of *samples* (see below).
In this latter case---perhaps, the one that more closely resembles decisions in people's daily life---the properties of the choice alternatives are not described and therefore not known with certainty, but they must be inferred from the consequences that occurred or were absent after the alternatives were chosen in similar decisions in the past.
Although these different formats may in principle carry the same information about given choice alternatives, behavioral decision research has yet produced a large body of papers indicating that, taking a quite general perspective, there are robust differences between the choices that are made in DfD and DfE.
That is, all other things being equal, empirical choice patterns across studies indicate that in DfD people choose *as if* a rare consequence of a choice alternative has more impact on the latter's evaluation than would be objectively warranted by the principle of mathematical expectation (see below); in turn, in DfE, people choose *as if* the same rare consequence has less impact on the evaluation than warranted [e.g., @barronSmallFeedbackbasedDecisions2003; @hertwigDecisionsExperienceEffect2004; @weberPredictingRiskSensitivity2004; @hauDescriptionexperienceGapRisky2008; @rakowBiasedSamplesNot2008; @ungemachAreProbabilitiesOverweighted2009; @camilleriWhenWhyRare2011; @kellenHowVariantAre2016; @erevAnomaliesForecastsDescriptive2017; @regenwetterConstructBehaviorGap2017; see also @wulffMetaanalyticReviewTwo2018, for a comprehensive meta-analytic review].

This paper uses the *description-experience gap* [@hertwigDescriptionexperienceGapRisky2009] as an entry point for a closer inquiry into the sampling process as an essential part of sample-based decision making.

<!--
The paper is structured as follows: 
...

All simulations and analyses are reproducible---materials and instructions can be found on the GitHub repository.
-->

@@

Now, more formally,^[Paragraphs with this beginning usually formalize the main ideas of the preceding paragraphs or elaborate on them by using mathematical expressions and concepts. Although important, the gist of this paper should be captured despite skipping these paragraphs.]  let the axiomatic definition of a prospect be that of a discrete random variable's distribution.
Let therefore the tuple $(\Omega, \mathcal{F}, P)$ be a probability space. 
Following the @kolmogorovFoundationsTheoryProbability1950 axioms, $\Omega$ is the sample space containing a finite set of elements
$$
\omega_i \in \{\omega_1, ..., \omega_n\} = \Omega
\; .
\tag{2}
$$ 
$\mathcal{F}$ is the set of all possible subsets of $\Omega$, i.e., the power set $\mathcal{P}(\Omega)$ with each element $A \in P(\Omega) = \mathcal{F}$ being a subset of $\Omega$.^[Note that this paper considers only the case of finite sample spaces. The exact axiomatic definition of $\mathcal{F}$, which can be found in Kolmogorov [-@kolmogorovFoundationsTheoryProbability1950, pp. 2-3 and 14-15], also covers infinite sample spaces and is that of a $\sigma$-algebra. For the finite case, however, $\mathcal{P}(\Omega)$ can be considered as such [@georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.1].] 
$P$ is a probability measure
$$
P: \mathcal{F} \to [0,1]
\; ,
\tag{3}
$$
which assigns each subset $A \in \mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.<!--^[The probabilities $p(\omega)$ are determined by the probability mass function $p : \Omega \to [0,1]$, which is used to construct the probability measures $P(A) = \sum_{\omega \in A} p(\omega)$ for all $A \in \mathcal{F}$.]-->
The random variable $X$ is the measurable function
$$
\begin{aligned}
  X : (\Omega, \mathcal{F}) &\to (\Omega', \mathcal{F'})  \\
  \omega &\mapsto x
  \; ,
\end{aligned}
\tag{4}
$$
where $(\Omega', \mathcal{F'})$ is a measurable space with each element $x \in \Omega'$ being a possible outcome of a prospect and each element $A' \in \mathcal{F'}$ being a subset of $\Omega'$.
An outcome's probability $p_X(x)$ is then provided by the probability measure
$$
P_X : \mathcal{F'} \to [0,1]
\tag{5}
$$
with 
$$
P_X(A' \in \mathcal{F'}) := P(X^{-1} \ A') = P(\{\omega: X(\omega) = x \in A'\})
\tag{6}
$$
[see @kolmogorovFoundationsTheoryProbability1950, p. 21; see also @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1, for a comprehensive introduction to random variables].
That is, the set consisting of a single outcome $x$ can be treated as a subset $\{x\} \in \mathcal{F'}$ and we therefore obtain $p_X(x) = P(X(\omega) = x)$.
Note that the probability measure in (5) and (6), respectively, is a discrete random variable's distribution and therefore a prospect.
To summarize and simplify, given a choice between $1, ..., k$ prospects, the inputs from the environment or memory should provide information about the set
$$
C = \{P_{X_1}, ... , P_{X_k}\}
\tag{7}
$$
with each element $c \in C$ being a distribution $P_X$ of a random variable defined on a different probability space.
Accordingly, each $c$ can itself be considered a set of outcome-probability pairs $(x_i, p_X(x_i)) \in \{(x_1, p_X(x_1)), ..., (x_n, p_X(x_n))\}$.

@@

Of course, the accuracy of such inferences depends, ceteris paribus, on the size of the samples, since from the laws of large numbers it follows that the relative frequency with which an outcome occurs in a sample converges to the outcome's objective probability only as the size of a sample increases [see @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 5].
However, if the choices that people actually make between prospects are described, robust differences can be found between the choices in DfD and DfE, and the size of samples in DfE provides just one and only a partial explanation for this *description-experience gap* [see @hertwigDescriptionexperienceGapRisky2009; @wulffMetaanalyticReviewTwo2018]. 
The following section briefly addresses how choices between prospects in general and the description-experience gap in particular can be described using cumulative prospect theory's [CPT, @tverskyAdvancesProspectTheory1992; @wakkerAxiomatizationCumulativeProspect1993] *weighting function*.^[This paper also considers CPT's value function, but emphasis is put on the weighting function.]
<!-- The two sections thereafter ... -->

# The Description of Choices in CPT and the Description-Experience Gap

@@

# Explanations for Underweighting in DfE

# Decision and Sampling Strategies

<!-- old -->

@@

# Evaluation of Choice Alternatives and Outputs of the Decision Making Process

# Sampling Strategies as Representation Systems

@@

## Test set

@@

## Results

@@

### Plausibility Check: Relationship between Switching Probability and Trial Length

@@

### Probability Weighting Function

@@

### Value Function

@@

# Discussion and Conclusion

>>>>

The human mind is a cognitive system that operates on inputs from an environment or memory, and in the domain of decision making, such inputs provide information about the *properties* of the choice alternatives between which people can choose.
In fact, irrespective of their exact mathematical specification or computational implementation, decision models generally take as inputs information about a set of choice alternatives to return a subset thereof as outputs [@heOntologyDecisionModels2020].

Now, more formally,^[Paragraphs with this beginning usually formalize the main ideas of the preceding paragraphs or elaborate on them by using mathematical expressions and concepts. Although important, the gist of this paper should be captured despite skipping these paragraphs.] let us treat the choice alternatives as elements $c$ of the set $C$.
To each $c \in C$ is then assigned some information from the inputs and this information may be used by the mind to learn about the properties of the respective alternative.
Accordingly, decision making can be considered an information-processing problem which starts from the set $C$, containing all choice alternatives, and stops at a proper subset $C' \subset C$ with the inclusion map
$$
\begin{aligned}
  \iota : C' &\to C \\
  c &\mapsto c  \quad \forall c \in C'
\end{aligned}
\tag{1}
$$
[@jostMathematicalConcepts2015, p. 15], containing only the chosen alternatives.

Importantly, the information that is provided about the choice alternatives can come in many forms.
In *decisions from description* (DfD), inputs take the form of explicit and complete descriptions of all relevant properties of the choice alternatives.
Yet, in *decisions from experience* [DfE, @hertwigDecisionsExperienceEffect2004], inputs take the form of *samples*.
In this latter case, the properties of the choice alternatives are not described and therefore not known with certainty, but they must be inferred from the outcomes that occurred when the alternatives were chosen in similar past decisions.
Although these distinct input forms may in principle carry the same information about a given set of choice alternatives, behavioral decision research has so far produced a large body of papers indicating that---if a quite general perspective is taken---there are robust differences between the choices that are made in DfD and DfE.
That is, ceteris paribus, empirical choice patterns across studies indicate that in DfD people choose *as if* a rare outcome is given more weight in the evaluation of a choice alternative than would be objectively warranted by the principle of mathematical expectation (see below); in turn, in DfE, people choose *as if* the same rare outcome is given less weight than would be warranted [e.g., @barronSmallFeedbackbasedDecisions2003; @hertwigDecisionsExperienceEffect2004; @weberPredictingRiskSensitivity2004; @hauDescriptionexperienceGapRisky2008; @rakowBiasedSamplesNot2008; @ungemachAreProbabilitiesOverweighted2009; @camilleriWhenWhyRare2011; @kellenHowVariantAre2016; @erevAnomaliesForecastsDescriptive2017; @regenwetterConstructBehaviorGap2017].
Hereafter, we refer to these patterns as the as-if overweighting and underweighting of rare outcomes, respectively.

This paper uses the choice pattern in DfE as an entry point for an inquiry into the sampling process as a consequential part of the information processing upfront choices in DfE.
More specifically, it will be discussed and demonstrated in a simulation how changes to the strategy of sequentially generating outcomes and integrating them into samples (hereafter: sampling strategies) can produce distinct choice patterns in DfE---among them, the as-if underweighting of rare outcomes [see @hillsInformationSearchDecisions2010, for the original formulation of this idea].

The rest of the paper is structured as follows.
In the first brief section, a formal definition of DfE is provided.
In the subsequent section, cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992; @wakkerAxiomatizationCumulativeProspect1993] is reviewed as a data model that can be used to describe choices in DfE in terms of deviations from the principle of mathematical expectation.
<!-- Outline of the section -->
In the section that follows, sampling strategies are treated in quite some detail.
<!-- Outline of the section -->
Thereafter, a simulation is presented in which the theoretical claims that are made in the preceding section are implemented in a computational model to simulate choice data in DfE.
The predictions of the model are summarized by fitting the choice data in CPT.
The paper closes with a summary and a conclusion.
Note that this paper, including the simulations and data analyses, is reproducible---materials and instructions can be found on the GitHub repository.

@@

Now, more formally, let the axiomatic definition of a prospect be that of a discrete random variable's distribution.
Let therefore the tuple $(\Omega, \mathcal{F}, P)$ be a probability space. 
Following the @kolmogorovFoundationsTheoryProbability1950 axioms, $\Omega$ is the sample space containing a finite set of elements
$$
\omega_i \in \{\omega_1, ..., \omega_n\} = \Omega
\; .
\tag{2}
$$ 
$\mathcal{F}$ is the set of all possible subsets of $\Omega$, i.e., the power set $\mathcal{P}(\Omega)$ with each element $A \in P(\Omega) = \mathcal{F}$ being a subset of $\Omega$.^[Note that this paper considers only the case of finite sample spaces. The exact axiomatic definition of $\mathcal{F}$, which can be found in Kolmogorov [-@kolmogorovFoundationsTheoryProbability1950, pp. 2-3 and 14-15], also covers infinite sample spaces and is that of a $\sigma$-algebra. For the finite case, however, $\mathcal{P}(\Omega)$ can be considered as such [@georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.1].] 
$P$ is a probability measure
$$
P: \mathcal{F} \to [0,1]
\; ,
\tag{3}
$$
which assigns each subset $A \in \mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.<!--^[The probabilities $p(\omega)$ are determined by the probability mass function $p : \Omega \to [0,1]$, which is used to construct the probability measures $P(A) = \sum_{\omega \in A} p(\omega)$ for all $A \in \mathcal{F}$.]-->
The random variable $X$ is the measurable function
$$
\begin{aligned}
  X : (\Omega, \mathcal{F}) &\to (\Omega', \mathcal{F'})  \\
  \omega &\mapsto x
  \; ,
\end{aligned}
\tag{4}
$$
where $(\Omega', \mathcal{F'})$ is a measurable space with each element $x \in \Omega'$ being a possible outcome of a prospect and each element $A' \in \mathcal{F'}$ being a subset of $\Omega'$.
An outcome's probability $p_X(x)$ is then provided by the probability measure
$$
P_X : \mathcal{F'} \to [0,1]
\tag{5}
$$
with 
$$
P_X(A' \in \mathcal{F'}) := P(X^{-1} \ A') = P(\{\omega: X(\omega) = x \in A'\})
\tag{6}
$$
[see @kolmogorovFoundationsTheoryProbability1950, p. 21; see also @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1, for a comprehensive introduction to random variables].
That is, the set consisting of a single outcome $x$ can be treated as a subset $\{x\} \in \mathcal{F'}$ and we therefore obtain $p_X(x) = P(X(\omega) = x)$.
Note that the probability measure in (5) and (6), respectively, is a discrete random variable's distribution and therefore a prospect.
To summarize and simplify, given a choice between $1, ..., k$ prospects, the inputs from the environment or memory should provide information about the set
$$
C = \{P_{X_1}, ... , P_{X_k}\}
\tag{7}
$$
with each element $c \in C$ being a distribution $P_X$ of a random variable defined on a different probability space.
Accordingly, each $c$ can itself be considered a set of outcome-probability pairs $(x_i, p_X(x_i)) \in \{(x_1, p_X(x_1)), ..., (x_n, p_X(x_n))\}$.

@@

Of course, the accuracy of such inferences depends, ceteris paribus, on the size of the samples, since from the laws of large numbers it follows that the relative frequency with which an outcome occurs in a sample converges to the outcome's objective probability only as the size of a sample increases [see @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 5].
However, if the choices that people actually make between prospects are described, robust differences can be found between the choices in DfD and DfE, and the size of samples in DfE provides just one and only a partial explanation for this *description-experience gap* [@hertwigDescriptionexperienceGapRisky2009; see also @wulffMetaanalyticReviewTwo2018, for a comprehensive meta-anayltic review]. 
The following section briefly addresses how choices between prospects in general and the description-experience gap in particular can be described using cumulative prospect theory's [CPT, @tverskyAdvancesProspectTheory1992; @wakkerAxiomatizationCumulativeProspect1993] *weighting function*.^[This paper also considers CPT's value function, but emphasis is put on the weighting function.]
<!-- The two sections thereafter ... -->

# The Description of Choices in CPT and As-If Underweighting in DfE

@@

# Sampling Strategies

<!-- Old

@@

-->

@@

<!-- Test set-->

@@

<!-- Results -->

@@

<!-- Plausibility Check: Relationship between Switching Probability and Trial Length -->

@@

<!-- Weighting function -->

@@

<!-- Weighting function -->

@@

# Summary and Conclusion

# Acknowledgement
